Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-24 numerical-amr.tex
edited section 3

Each grid patch in \enzo\ contains arrays of values for baryon and particle quantities.   Grids are partitioned into a core of \emph{active zones} and a surrounding layer of \emph{ghost zones}, as shown in Figure~\ref{fig.amr_hierarchy}.  The active zones store field values and ghost zones are used to temporarily store values that have been obtained directly from neighboring grids or interpolated from a parent grid.  These zones are necessary to accommodate the computational stencil of the (magneto)hydrodynamics solvers (Sections~\ref{sec.hydro.ppm} and~\ref{sec.hydro.zeus}) and the gravity solver (Section~\ref{sec.gravity}).  The PPM and Zeus hydro solvers require 3 layers of ghost zones,  and the gravity solver requires 6 (although only for the density and potential fields).  This can lead to significant memory and computational overhead, particularly for smaller grid patches at high levels of refinement.

@@

Timesteps are determined on each level as described in Section~\ref{sec.timestepping}, and the hierarchy is advanced on
a level-by-level basis in a W-Cycle.    Beginning with the coarsest level, $l$, all grids on that level are advanced one
timestep.  Then one timestep is taken on all grids at the next level of refinement, $l+1$, and so on until the finest level is resolved. The finest level is then advanced, using as many steps as it takes to reach the level immediately above.  The finest level is then synched to the level above it, which then proceeds forward in time one more step.  This is shown graphically in Figure~\ref{fig:wcycle}.

@@

The basic control algorithm looks very much like any used in a grid-based code of a single resolution, and can be written schematically as:
\begin{verbatim}
InitializeHierarchy
While (Time < StopTime)
   dt = ComputeTimeStep(0)
   EvolveLevel(0, dt)
   Time = Time + dt
  CheckForOutput(Time)
\end{verbatim}
The AMR control algorithm is contained within the recursive
EvolveLevel algorithm, which takes the level on which to operate as an
argument.  This looks something like the following pseudo-code:
\begin{verbatim}
EvolveLevel(level)
   SetBoundaryValues
   while (Time < ParentTime)
      dt = ComputeTimeStep(level)
      SolveHydroEquations(dt)
      SolveOtherEquations(dt)
      Time = Time + dt
      SetBoundaryValues
      EvolveLevel(level+1, dt)
      FluxCorrection
      Projection
      RebuildHierarchy(level+1)
\end{verbatim}

@@

In one dimension, such functions are well known; examples include
donor cell, Van Leer and piecewise parabolic interpolation, which are
first, second and third-order accurate respectively
\citep[e.g.,][]{Stone92a}.  However, we require fully
multi-dimensional interpolation functions, and so focus on
second-order schemes since they are tractable and relatively 
accurate\footnote{We could, in principle, build a multi-dimensional
  interpolator out of a one-dimensional interpolation function;
  however, at higher order, this would not necessarily satisfy our
  constraint regarding not introducing new extrema in interpolated quantities.}.  In Appendix~\ref{app:interpolation}, we
explicitly write down functions in one, two and three dimensions, but
here we discuss the general design considerations.

@@

In three dimensions this process breaks down since the number of
constraints (center + four diagonals = five) exceeds the number of
parameters (four).  One way to solve this problem would be to perform
a least-squares fit under the given constraints, however, this is
too computationally expensive.  Therefore, in Appendix~\ref{app:interpolation} we describe
a heuristic approach that satisfies four constaints analytically and
then adjusts the interpolation parameters in such a way as to satisfy the fifth.
While the result is not necessarily optimal, it does appear to perform
adequately in our tests.

@@

\item{\em Jeans length.}  Following \cite{Truelove98}, the code can be forced to refine the Jeans length by a fixed number of cells, or more specifically, whenever the following criterion is not met:
\begin{equation}
\Delta x < \left( \frac{\pi k_B T}{J^2 G \rho \mh} \right)^{1/2},
\end{equation}
where $J$ is the required number of cells per Jeans length (4 by default).

\item{\em Cooling time.}  Often in cosmological simulations, cooling is a rapid process and we do not necessarily want to be limited by the cooling time.  However, in some cases it is desirable to resolve the cooling process, in which case we flag cells that have a cooling time shorter than their sound crossing time.  The cooling time is given by $t_{\rm cool} = 1.5 kT / (n\, \Lambda(T))$ and the sound crossing time is $t_{\rm cross} = \Delta x / c_s$, where $c_s$ is the sound speed.

@@

\item{\em Resistive length.}  In a way similar to the Jeans length criterion described above, the code can be forced to locally resolve the resistive length scale by a minimum of a fixed number of cells.  This is done by ensuring that the resistive length scale, defined as
\begin{equation}
L_R = |B| / |\curl B|
\end{equation}
is always refined by at least N cells locally (i.e.~we refine a given cell further if $L_R \leq N \Delta x$).  Note that \enzo\ does not actually solve the resistive MHD equations -- this is the ``characteristic'' local resistive length, rather than one calculated within the MHD solver.

@@

Finally, it is also possible to specify a rectangular region that is statically refined to a certain level.  These static refined regions are in place throughout the simulation regardless of the grid quantities and can be used for testing purposes, or for multiply-zoomed initial conditions (for example in cosmology simulations) that would not otherwise be refined.

>>>>

Each grid patch in \enzo\ contains arrays of values for baryon and particle quantities.   Grids are partitioned into a core of \emph{active zones} and a surrounding layer of \emph{ghost zones}, as shown in Figure~\ref{fig.amr_hierarchy}.  The active zones store field values and ghost zones are used to temporarily store values that have been obtained directly from neighboring grids or interpolated from a parent grid.  These zones are necessary to accommodate the computational stencil of the (magneto)hydrodynamics solvers (Sections~\ref{sec.hydro.ppm} through~\ref{sec.hydro.zeus}) and the gravity solver (Section~\ref{sec.gravity}).  The PPM and Zeus hydro solvers require 3 layers of ghost zones,  and the gravity solver requires 6 (although only for the density and potential fields).  This can lead to significant memory and computational overhead, particularly for smaller grid patches at high levels of refinement.

@@

Timesteps are determined on each level as described in Section~\ref{sec.timestepping}, and the hierarchy is advanced on
a level-by-level basis in a W-Cycle.    Beginning with the coarsest level, $l$, all grids on that level are advanced one
timestep.  Then, one timestep is taken on all grids at the next level of refinement, $l+1$, and so on until the finest level is resolved. The finest level is then advanced, using as many steps as it takes to reach the level immediately above.  The finest level is then synched to the level above it, which then proceeds forward in time one more step.  This is shown graphically in Figure~\ref{fig:wcycle}.

@@

The basic control algorithm looks very much like any used in a grid-based code of a single resolution, and can be written schematically as:
\begin{verbatim}
InitializeHierarchy
While (Time < StopTime)
   dt = ComputeTimeStep(0)
   EvolveLevel(0, dt)
   Time = Time + dt
   CheckForOutput(Time)
\end{verbatim}
The AMR control algorithm is contained within the recursive
EvolveLevel algorithm, which takes the level upon which to operate as an
argument.  This looks something like the following pseudo-code:
\begin{verbatim}
EvolveLevel(level)
   SetBoundaryValues
   while (Time < ParentTime)
      dt = ComputeTimeStep(level)
      SolveHydroEquations(dt)
      SolveOtherEquations(dt)
      Time = Time + dt
      SetBoundaryValues
      EvolveLevel(level+1, dt)
      FluxCorrection
      Projection
      RebuildHierarchy(level+1)
\end{verbatim}

@@

In one dimension, such functions are well known; examples include
donor cell, Van Leer, and piecewise parabolic interpolation, which are
first-, second-, and third-order accurate respectively
\citep[e.g.,][]{Stone92a}.  However, we require fully
multi-dimensional interpolation functions, and so focus on
second-order schemes since they are tractable and relatively 
accurate\footnote{We could, in principle, build a multi-dimensional
  interpolator out of a one-dimensional interpolation function;
  however, at higher order, this would not necessarily satisfy our
  constraint regarding not introducing new extrema in interpolated quantities.}.  In Appendix~\ref{app:interpolation}, we
explicitly write down functions in one, two and three dimensions, but
here we discuss the general design considerations.

@@

In three dimensions this process breaks down since the number of
constraints (center + four diagonals = five) exceeds the number of
parameters (four).  One way to solve this problem would be to perform
a least-squares fit under the given constraints; however, this is
too computationally expensive.  Therefore, in Appendix~\ref{app:interpolation} we describe
a heuristic approach that satisfies four constaints analytically and
then adjusts the interpolation parameters in such a way as to satisfy the fifth.
While the result is not necessarily optimal, it does appear to perform
adequately in our tests.

@@

\item{\em Jeans length.}  Following \cite{Truelove98}, the code can be forced to refine the Jeans length by a fixed number of cells, or, more specifically, whenever the following criterion is not met:
\begin{equation}
\Delta x < \left( \frac{\pi k_B T}{N_J^2 G \rho \mh} \right)^{1/2},
\end{equation}
where $J_J$ is the required number of cells per Jeans length (4 by default).

\item{\em Cooling time.}  Often in cosmological simulations, cooling is a rapid process and we do not necessarily want to be limited by the cooling time.  However, in some cases it is desirable to resolve the cooling process, in which case we flag cells that have a cooling time shorter than their sound crossing time.  The cooling time is given by $t_{\rm cool} = 1.5 k_BT / (n\, \Lambda(T))$ and the sound crossing time is $t_{\rm cross} = \Delta x / c_s$, where $c_s$ is the sound speed.

@@

\item{\em Resistive length.}  In a way similar to the Jeans length criterion described above, the code can be forced to locally resolve the resistive length scale by a minimum of a fixed number of cells.  This is done by ensuring that the resistive length scale, defined as
\begin{equation}
L_R = |B| / |\curl B|
\end{equation}
is always refined by at least N$_R$ cells locally (i.e.~we refine a given cell further if $L_R \leq N_R \Delta x$).  Note that \enzo\ does not actually solve the resistive MHD equations -- this is the ``characteristic'' local resistive length, rather than one calculated within the MHD solver.

@@

Finally, it is also possible to specify a rectangular region that is statically refined to a certain level.  These static refined regions are in place throughout the simulation regardless of the grid quantities and can be used for testing purposes, or for multiply-zoomed initial conditions (in cosmology simulations, for example) that would not otherwise be refined.
